Data exploration, quality control and statistical analyses of ChIP-exo experiments\vspace*{\fill}

### 2015-11-27 rmarkdown/prelims_paper.tex
Second version of the paper finished

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that background distribution of ChIP-exo data is less
homogeneous than that of ChIP-Seq data.

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/forward_strand_ratio_comp_old.pdf}
\includegraphics[width = .46\textwidth,page = 1]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \includegraphics[width = .46\textwidth,page = 2]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there were clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty. B) In ChIP-exo data, strands of reads were significantly
    less balanced in the regions with potential binding sites compared
    to SET ChIP-Seq data. C) ChIP tag counts increase linearly as
    mappability scores increases. D) ChIP tag counts increase linearly
    as GC content score increases when GC content is less than 0.6 and
    then ChIP tag counts decrease as GC content increases. }
  \label{fig:comp}
\end{figure}

@@

Using the mouse-FoxA1 experiment from \cite{exoillumina} and the
relationship between this two quantities, library complexity and
sample enrichment was explored. Figure \ref{fig:enrich}A shows hexbin
plots illustrating the interaction between these two
quantities. Notice how there are two strong arms in each panel: The
first one corresponds to regions with low $\mbox{ARC}$ values and
varying $\mbox{URCR}$ values across the $(0,1]$ interval, while the
second one shows a decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$
increases. When an experiment shows a higher degree of enrichment,
then the separation of this two arms is more noticeable, since the
second arm corresponds to possibly enriched regions
(\ref{fig:enrich2}A and \ref{fig:enrich2}B).

Figure \ref{fig:enrich}B shows box plots of the $\mbox{local-NSC}$ on
regions of the three replicates, stratified by the number of reads
mapped exactly to one position for all $\sigma^{70}$ ChIP-exo
experiments; The high stratum is defined by regions consisting of more
than 100 unique positions, the medium stratum for regions
where the number of unique position is on the
(50,100) range and the low stratum in the
(20,50) range. For each stratum and biological
replicate, we sampled 400 regions whenever possible (in the
opposite case, all the regions in the category were considered) and
calculated the $\mbox{local-NSC}$ for those regions. For all three
replicates, it is shown an increasing trend as the number of unique
position increases, which means that the $\mbox{local-NSC}$ is an
effective indicator of library complexity. Figure \ref{fig:enrich}C
shows the $\mbox{local-NSC}$ coefficient calculated for the same set
of regions, stratified by the number of unique positions in the
regions of the first replicate using the same criteria as before. In
the high panel it is shown that the first replicate sample performs
slightly better than the second and both outperform the third one,
this agrees with \ref{fig:enrich}A, where only in the left-most panel
doesn't seem to be a separation between the low $\mbox{ARC}$ arm and
the other.

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .7\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
%  \includegraphics[width = .7\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate.}
  \label{fig:enrich}
\end{figure}

@@

The strand imbalance assessment is based in the observation that the
enriched regions usually have a higher concentration of fragments,
therefore we examined the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. However, figure \ref{fig:comp}B shows that this value is
not representative of the sample locally, therefore the assumption
doesn't hold in practice.

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least 100
fragments into into the ones that overlap with any of the sample's
peaks. Then we considered a Kolmogorov-Smirnov tests to assess if the
distribution of both groups were the same. The p.value for the third
replicate's test is 0.1039, for the first replicate's test
is 0.0064 and for the second test is
0.0871. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the 100
level.

\begin{figure}[H]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in figure
    \ref{fig:enrich}. A) FSR distribution quantiles as the lower depth
    regions are being filtered out, all quantiles approach to the
    median as the lower bound increases. B) Stacked histogram with the
    proportion of regions that are formed by two strands or only one,
    in a good sample the single-stranded regions are going to be
    filtered out quickly as in the middle row.}
  \label{fig:strand}
\end{figure}

@@

Figure \ref{fig:reso_all} shows different comparisons among ChIP-exo,
PET ChIP-Seq and SET ChIP-Seq. A RegulonDB annotation was considered
identified if the distance between it and dPeak binding site estimate
was at most of 20 bp. That way, the sensitivity is defined as
the proportion of RegulonDB annotations identified in a peak. In
figure \ref{fig:reso_all}A can be seen that the sensitivity of all
protocols increases as the average distance between binding sites
does. Despite that when the binding events in a peak are closer to
each other, both ChIP-exo and PET ChIP-Seq are comparable, as the
distance increases ChIP-exo identifies a higher proportion of the
RegulonDB annotations; additionally SET ChIP-Seq is significantly less
sensible than both ChIP-exo and PET ChIP-Seq. In figure
\ref{fig:reso_all}B the distance between a RegulonDB annotation to its
closest prediction is compared for ChIP-exo, PET and SET ChIP-Seq;
while the first two are comparable, both outperform SET ChIP-Seq.

In figures \ref{fig:reso_all}C and \ref{fig:reso_all}D, we observe the
behavior of dPeak's estimated parameters for single end data (both
ChIP-exo and ChIP-Seq). The $\delta$ parameter measures the distance
from the 5' end of the fragment to its respective binding event and
the $\sigma$ parameter indicates the fragments distribution
variability around their respective binding sites (more information in
\cite{dpeak}). For both parameters, it can be seen that the ChIP-exo
estimates are smaller than the SET ChIP-Seq estimates in average,
which agrees with the fact that the sample reads are allocated more
tightly around the binding events in ChIP-exo data. Hence we can
conclude that the peaks shape is quite different between ChIP-exo and
SET ChIP-Seq.

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth]{../figs/for_paper/sensitivity_exo_olda_data.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/resolution_by_dataset_old_data.pdf}
\end{figure}

@@

   \caption{Comparison of (A) sensitivity and (B) resolution between
     ChIP-exo and ChIP-Seq data. Sensitivity is defined as the
     proportion of RegulonDB annotations identified using each
     data. Resolution is defined as the distance between RegulonDB
     annotation and its closest prediction. $\quad$ C) $\delta$ parameter in
     dPeak measures average distance of the reads to their respective
     binding site. In ChIP-exo data, reads were located much closer to
     the binding site than in SET ChIP-Seq. D) $\sigma$ parameter
     measure the dispersion of reads around each binding site. In
     ChIP-exo data, reads showed less variation around the their
     respective binding sites compared to SET ChIP-Seq.}
  \label{fig:reso_all}
\end{figure}

@@

We sampled a fixed amount of fragments for each of the ChIP-exo, PET
ChIP-Seq and SET ChIP-Seq datasets of the $\sigma^{70}$ sample in
aerobic conditions. For each sampled dataset we applied our
lower-to-higher resolution pipeline by calling peaks with MOSAiCS
\cite{mosaics} and then deconvolving the binding events by using dPeak
\cite{dpeak}. For the ChIP-exo datasets we called peaks by using
GC-content and mappability with MOSAiCS, and for the ChIP-Seq datasets
we used their respective Input samples.

\begin{figure}[H]
  \centering
%  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_aerobic_saturation.pdf}
  \includegraphics[width = .8\textwidth]{../figs/for_paper/saturation_analysis_old.pdf}
  \caption{Comparison of the number of candidate regions (A),
    predicted events (B), identified targets (C) and resolution (D)
    among ChIP-exo, PET ChIP-Seq and SET ChIP-Seq. RegulonDB
    annotations are considered as a gold standard. A gold standard was
    considered identified if a BS was estimated inside a 15
    bp window around it}
  \label{fig:design}
\end{figure}

Figure \ref{fig:design} shows the behavior of each data type when
their depth is fixed. It is remarkable that even when the number of
candidate peaks or the number of predicted events is lower for
ChIP-exo, it outperforms both PET and SET ChIP-Seq in the number of
identified targets and resolution. This may suggest that with ChIP-exo
less false positive peaks are being called and that when the targets
are being identified, dPeak estimates binding locations closer to the
true location. Additionally, we can see that as the read depth
increases all four indicators do so as well, which may indicate that
with ChIP-exo a smaller amount of reads is necessary to identify a
higher number of targets, but it may also be possible that this is an
artifact occurring due to ChIP-exo's lower library complexity.

@@

\begin{itemize}
\item \textbf{dPeak}: We updated the initialization strategy. The
  latest version is currently available from
  \url{http://dongjunchung.github.io/dpeak/}.
\item \textbf{ChIPexoQual}: This package contains the QC pipeline for
  ChIP-exo. The last version is available in
  \url{https://github.com/welch16/ChIPexoQual}.
\item \textbf{Segvis}: The goal of this package is to visualize
  genomic regions by using aligned reads such as in figures
  \ref{fig:exo} or \ref{fig:qcdiagram}.  The latest version is
  available in \url{https://github.com/keleslab/Segvis}.
\item \textbf{ChIPUtils}: This package attempts to gather the most
  commonly used ChIP-Seq QC (as in table \ref{tab:qcbase} or figure
  \ref{fig:scc}) indicators and additional utilities (as figure
  \ref{fig:comp}A) for ChIP-Seq. The latest available version is in
  \url{https://github.com/welch16/ChIPUtils}.
\end{itemize}

@@

We provide a ChIP-exo QC pipeline capable of assess the balance between
enriched samples and low complexity regions. It is shown that the
``peak-pair'' assumption doesn't hold in practice and we provide two
out-of-the-box visualization capable to assess the strand imbalance in
a ChIP-exo experiment.

We updated the dPeak algorithm from \cite{dpeak} and showed that
ChIP-exo is comparable in resolution and sensitivity with PET ChIP-Seq
and outperforms SET ChIP-Seq. ChIP-exo compared with PET and SET
ChIP-Seq at fixed depth sample is able to identify more targets at a
lower resolution. We compared dPeak with another algorithms to
estimate binding locations in ChIPexo data, dPeak is comparable to
MACE and outperforms GEM in resolution. dPeak provided a striking
balance in sensitivity, specificity and spatial resolution for
ChIP-exo analysis.

@@

Figure \ref{fig:comp} shows a relationship between the ChIP tag counts
of a ChIP-exo experiment and both mappability and GC content
scores. Furthermore, in sections \ref{sec:reco} and
\ref{sec:dpeak_analysis} we called peaks by using MOSAiCS which
considers them as necessary information when there is not input
sample. Hence, quality measures we may consider to add indicators to
the QC pipeline that considers this biases.

@@

He et al., 2014 \cite{chipnexus} proposed a modification to the
ChIP-exo protocol that may fix its low library complexity issues by
extending the fragments by a random barcode prior to the exonuclease
digestion. This last statement is briefly mentioned by Mahony and
Pugh, 2015 \cite{exo_review}. We want to apply our QC pipeline to
ChIP-Nexus. Depending if ChIP-Nexus data shows background we may need
to modify the QC pipeline in the step where it partitions the reads
into regions, but the $\mbox{local-NSC}$ may be used to prove this
statement.

@@

The data showed in figure \ref{fig:exo_example} is actually part of a
more interesting experiment. We have 2 biological replicates of the
$\sig$, $\beta'_f$ and $\beta$ transcription factors under two
conditions, one where rifampicin was applied by 20 minutes and another
where it wasn't applied. As seen in the figure, there is ChIP-exo data
for the 3 factors and PET ChIP-Seq for $\sig$ and $\beta'_f$ (in the
figure both replicates were pooled).

$\sig$ factor is a transcription initiation factor of housekeeping
genes in E. Coli, and both $\beta$ factors transcribe the DNA into
mRNA. Hence, we want to use this data to have a better understanding
of the transcription in E. Coli.

@@

\begin{figure}[H]
  \centering
 \includegraphics[width = \textwidth]{/p/keles/ChIPexo/volume5/OpenClosedComplexes/figs/filtered/heatmaps_centroid.png}
 \caption{Centered and normalized heatmaps for PET ChIP-Seq peaks of
   $\sig$ and $\beta'_f$ under rif0 and rif20. The regions are
   organized by the centroids clustering of the $\beta'_f$-rif0
   profiles.}
  \label{fig:hm1}
\end{figure}

Figure \ref{fig:hm1} shows heatmaps of the normalized coverages of PET
ChIP-Seq peaks for $\beta'_f$ and $\sig70$ under rif0 and rif20
treatments. Each row represents the same region, under the four
different samples. As expected the signal for both $\sig$ samples is
centered around the middle, for $\beta'_f$-rif20, the majority of he
signal seems to be centered, hence the transcription was interrupted
and for $\beta'_f$-rif0 we can see clusters formed by open regions.

Then, using a set of high quality PET ChIP-Seq peaks and clustering
with the mclust model by Fraley and Raftery, 2007 \cite{mclust} we
were able to label the a set of regions as closed or open
complexes. Hence, we expect that by using ChIP-exo data we may find
different patterns that represent transcription units whose output was
limited at different steps in the formation of EC$_2$.

@@

Last summer, we entered to a contest from ENCODE's functional
characterization group to fit a model to classify enhancers. Roughly
speaking, the dataset consisted of a set of regions
$\mathcal{R}_1,\cdots,\mathcal{R}_n$ and a set of labels
$Y_1,\cdots,Y_n \in \{0,1,\}$. Hence, we built a series of features
$X_1,\cdots,X_p$ and fitted a model by dividing the $n$ observations
into a training set and a test set, etc. However, there are in total
another $m >> n$ regions in the experiment, and the cost of labeling
those $m$ regions may be high (in either the actual monetary cost of
the experiment or the time the scientists spends labeling those
regions). Those $n+m$ regions are H3K27ac peaks, hence they were
ranked and the first $n$ were selected randomly from groups with high,
medium and low signal. This raises the following questions:

>>>>

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/forward_strand_ratio_comp_old.pdf}
\includegraphics[width = .46\textwidth,page = 1]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \includegraphics[width = .46\textwidth,page = 2]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \caption{ A) Hexbin plot of PE ChIP-Seq bin counts VS ChIP-exo bin
    counts. B) Forward Strand Ratio densities for SE ChIP-Seq and
    ChIP-exo peaks. C) ChIP tag counts increase linearly as
    mappability scores increases. D) ChIP tag counts increase linearly
    as GC content score increases when GC content is less than 0.6 and
    then ChIP tag counts decrease as GC content increases. }
  \label{fig:comp}
\end{figure}

@@

Figure \ref{fig:enrich}A shows hexbin plots illustrating the
relationship between library complexity and sample enrichment for the
FoxA1 in mouse cells from \cite{exoillumina}. Notice how there are two
strong arms in each panel: The first one corresponds to regions with
low $\mbox{ARC}$ values and varying $\mbox{URCR}$ values across the
$(0,1]$ interval, while the second one shows a decreasing trend in
$\mbox{URCR}$ as $\mbox{ARC}$ increases. When an experiment shows a
higher degree of enrichment, then the separation of this two arms is
more noticeable, since as Figures \ref{fig:enrich2}A and
\ref{fig:enrich2}B show the ARC VS URCR plots but filtering out the
regions formed by a low number of unique positions (10 and 30
respectively), hence this second arm corresponds to possibly enriched
regions.

Figure \ref{fig:enrich}B shows box plots of the $\mbox{local-NSC}$ on
regions of the three replicates, stratified by the number of reads
mapped exactly to one position for all Fox-A1 ChIP-exo experiments;
The high stratum is defined by regions consisting of more than
100 unique positions, the medium stratum for regions where
the number of unique position is on the (50,100)
range and the low stratum in the (20,50) range. For
each stratum and biological replicate, we sampled 400
regions whenever possible (in the opposite case, all the regions in
the category were considered) and calculated the $\mbox{local-NSC}$
for those regions. On one side, we can see that for all three
replicates, as the number of unique positions increases, then the
$\mbox{local-NSC}$ shows higher values. In the low panel, all three
replicates are comparable, in the medium panel, the first two
replicates are comparable and both outperform the third one, and
finally in the high panel we can see that despite that the first two
replicates are comparable, the first one slightly outperforms the
second. We can see that all three panels coincide with the conclusion
we get from Figure \ref{fig:enrich}A: First, the first replicate seems
to have the highest library complexity and the third replicate the
lowest. Second, despite that both first two replicates show a similar
level of bottlenecking, the regions in the first replicate seems to
have a higher ARC, those regions seem to be contributing to have a
higher local-NSC in the high panel. Third, despite that the ARC of the
first and third replicates is comparable, the third replicate has a
higher bottlenecking level which cause to $\mbox{local-NSC}$ to have
lower levels overall.

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .7\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
%  \includegraphics[width = .7\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate.}
  \label{fig:enrich}
\end{figure}

@@

The strand imbalance assessment is based in the observation that the
enriched regions usually have a higher concentration of reads,
therefore we examined the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In Table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. However, Figure \ref{fig:comp}B shows that this value
doesn't represent the sample locally, hence the assumption doesn't
hold well in practice.

In order to assess the strand imbalance we created the following
visualization presented in Figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by reads of both strands
or only one (forward or backward). As the number of reads per region's
lower bound increases, the quantiles tend to approach the median. To
evaluate this statement, for each replicate of FoxA1, we divided the
regions formed by at least 100 reads into the ones that overlap
with any of the sample's peaks. Then we considered Kolmogorov-Smirnov
tests to assess if the FSR's distribution was the same for the islands
that overlap with a peak and the islands that didn't. The p.value for
the third replicate's test is 0.1039, for the first
replicate's test is 0.0064 and for the second test is
0.0871. This shows that only the middle sample from Figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the 100
level.

\begin{figure}[H]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in Figure
    \ref{fig:enrich}. A) FSR distribution quantiles as the lower depth
    regions are being filtered out, all quantiles approach to the
    median as the lower bound increases. B) Stacked histogram with the
    proportion of regions that are formed by two strands or only one,
    in a good sample the single-stranded regions are going to be
    filtered out quickly as in the middle row.}
  \label{fig:strand}
\end{figure}

@@

Figure \ref{fig:reso_all} shows different comparisons among ChIP-exo,
PE ChIP-Seq and SE ChIP-Seq. A RegulonDB annotation was considered
identified if the distance between it and dPeak binding site estimate
was at most of 20 bp. That way, the sensitivity is defined as
the proportion of RegulonDB annotations identified in a peak (Salgado
et al., 2012 \cite{regulonDB}). Figure \ref{fig:reso_all}A illustrates
that the sensitivity of all protocols increases as the average
distance between binding sites does. Despite that when the binding
events in a peak are closer to each other, both ChIP-exo and PE
ChIP-Seq are comparable, as the distance increases ChIP-exo identifies
a higher proportion of the RegulonDB annotations; additionally SE
ChIP-Seq is significantly less sensitive than both ChIP-exo and PE
ChIP-Seq. In Figure \ref{fig:reso_all}B the distance between a
RegulonDB annotation to its closest prediction is compared for
ChIP-exo, PE and SE ChIP-Seq; while the first two are comparable,
both outperform SE ChIP-Seq.

In Figures \ref{fig:reso_all}C and \ref{fig:reso_all}D, we observe the
behavior of dPeak's estimated parameters for single end data (both
ChIP-exo and ChIP-Seq). The $\delta$ parameter measures the distance
from the 5' end of the read's 5' end to its respective binding event
and the $\sigma$ parameter indicates the read's starting position
distribution variability around their respective binding sites (more
information in \cite{dpeak}). For both parameters, it can be seen that
the ChIP-exo estimates are smaller than the SE ChIP-Seq estimates in
average, which agrees with the fact that the sample reads are
allocated more tightly around the binding events in ChIP-exo
data. This confirms that the dPeak model is able to recover the same
conclusions that have been done when comparing ChIP-exo and SE
ChIP-Seq.

% Hence we can conclude that the peaks shape is quite different
% between ChIP-exo and SE ChIP-Seq.
\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth]{../figs/for_paper/sensitivity_exo_olda_data.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/resolution_by_dataset_old_data.pdf}
\end{figure}

@@

   \caption{Comparison of (A) sensitivity and (B) resolution between
     ChIP-exo and ChIP-Seq data. Sensitivity is defined as the
     proportion of RegulonDB annotations identified using each
     data. Resolution is defined as the distance between RegulonDB
     annotation and its closest prediction. $\quad$ C) $\delta$ parameter in
     dPeak measures average distance of the reads to their respective
     binding site. In ChIP-exo data, reads were located much closer to
     the binding site than in SE ChIP-Seq. D) $\sigma$ parameter
     measure the dispersion of reads around each binding site. In
     ChIP-exo data, reads showed less variation around the their
     respective binding sites compared to SE ChIP-Seq.}
  \label{fig:reso_all}
\end{figure}

@@

Previously, ChIP-exo and SE ChIP-Seq have been compared at a fixed
depth level in the literature, but this comparisons didn't included PE
ChIP-Seq. Hence, we sampled a fixed amount of reads for each of the
ChIP-exo, PE ChIP-Seq and SE ChIP-Seq datasets of the $\sigma^{70}$
sample in aerobic conditions. For each sampled dataset we applied our
lower-to-higher resolution pipeline by calling peaks with MOSAiCS
\cite{mosaics} and then deconvolving the binding events by using dPeak
\cite{dpeak}. For the ChIP-exo datasets we called peaks by using
GC-content and mappability with MOSAiCS, and for the ChIP-Seq datasets
we used their respective Input samples.

\begin{figure}[H]
  \centering
%  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_aerobic_saturation.pdf}
  \includegraphics[width = .8\textwidth]{../figs/for_paper/saturation_analysis_old.pdf}
  \caption{\textbf{ChIP-exo, PE ChIP-Seq and SE ChIP-Seq comparison at
      varying sequencing depths}Comparison of the number of candidate
    regions (A), predicted events (B), identified targets (C) and
    resolution (D) among ChIP-exo, PE ChIP-Seq and SE
    ChIP-Seq. RegulonDB annotations are considered as a gold
    standard. A gold standard binding events was deemed identified if
    a binding event was estimated at a $\pm$ 15 vicinity of
    it.}
  \label{fig:design}
\end{figure}

Figure \ref{fig:design} shows the behavior of each data type when
their depth is fixed. It is remarkable that even when the number of
candidate peaks or the number of predicted events is lower for
ChIP-exo, it outperforms both PE and SE ChIP-Seq in the number of
identified targets and resolution. This may suggest that with ChIP-exo
less false positive peaks are being called and that when the targets
are being identified, dPeak estimates binding locations closer to the
true location. Additionally, we can see that as the read depth
increases all four indicators do so as well, which may indicate that
with ChIP-exo a smaller amount of reads is necessary to identify a
higher number of targets, but it may also be possible that this is an
artifact occurring due to ChIP-exo's lower library
complexity. Additionally, it is worth noting that for PE ChIP we
sampled both ends of the fragment, hence for each sequencing depth
considered we are sampling twice as many reads for PE ChIP-Seq than
for ChIP-exo or SE ChIP-seq.

@@

\begin{itemize}
\item \textbf{dPeak}: We updated the initialization strategy. The
  latest version is currently available from
  \url{http://dongjunchung.github.io/dpeak/}.
\item \textbf{ChIPexoQual}: This package contains the QC pipeline for
  ChIP-exo. The last version is available in
  \url{https://github.com/welch16/ChIPexoQual}.
\item \textbf{Segvis}: The goal of this package is to visualize
  genomic regions by using aligned reads such as in Figures
  \ref{fig:exo} or \ref{fig:qcdiagram}.  The latest version is
  available in \url{https://github.com/keleslab/Segvis}.
\item \textbf{ChIPUtils}: This package attempts to gather the most
  commonly used ChIP-Seq QC (as in Table \ref{tab:qcbase} or Figure
  \ref{fig:scc}) indicators and additional utilities (as Figure
  \ref{fig:comp}A) for ChIP-Seq. The latest available version is in
  \url{https://github.com/welch16/ChIPUtils}.
\end{itemize}

@@

We provide a ChIP-exo QC pipeline capable of assess the balance
between enriched samples and low complexity regions. It is shown that
the ``peak-pair'' assumption doesn't hold locally in practice and we
provide two out-of-the-box visualization capable to assess the strand
imbalance in a ChIP-exo experiment.

We showed that ChIP-exo is comparable in resolution with PE ChIP-Seq
and both outperforms SE ChIP-Seq. ChIP-exo shows higher sensitivity
than ChIP-Seq (both PE and SET), and this effect increases when the
average distance between binding events is larger.

We demonstrated for the first time that when ChIP-exo is compared with
PE ChIP-Seq (and SE ChIP-Seq) at a fixed sequencing depth, ChIP-exo is
able to identify more targets at a lower resolution. dPeak provides a
striking balance in sensitivity, specificity and spatial resolution
for ChIP-exo analysis.

% We compared dPeak with another
% algorithms to estimate binding locations in ChIP-exo data, dPeak is
% comparable to MACE and outperforms GEM in resolution. dPeak provided a
% striking balance in sensitivity, specificity and spatial resolution
% for ChIP-exo analysis.

@@

Figure \ref{fig:comp} shows a relationship between the ChIP tag counts
of a ChIP-exo experiment and both mappability and GC content
scores. Furthermore, in sections \ref{sec:reco} and
\ref{sec:dpeak_analysis} we called peaks by using MOSAiCS which
considers them as necessary information when there is not control (the
GOF plot is in Figure \ref{fig:mosaics_exo}) sample. Hence, we may add
QC measures that consider the GC content and mappability score biases
into the QC pipeline.

@@

% For example, the recently described ChIP-nexus protocol ligates both
% sequencing adaptors onto one end of ChIP fragments in a single step
% (as opposed to two separate ligation steps in the original ChIP-exo
% protocol) (He et al., 2015). Exonuclease digestion, DNA
% self-circularization with circLigase, and restriction enzyme cutting
% between the two adaptors creates the final library. By removing one
% of the relatively inefficient ligation steps, ChIP-nexus yields
% higher complexity libraries with the same resolution as
% ChIP-exo. One concern is that circLigase activity might have
% sequence specificity (Kwok et al., 2013), potentially creating bias
% in the identification of binding locations.

He et al., 2014 \cite{chipnexus} proposed a modification to the
ChIP-exo protocol that may fix its low library complexity issues by
extending the reads by a random barcode prior to the exonuclease
digestion and ligating both sequencing adaptors onto one end of the
ChIP DNA fragment (as opposed to two separate ligation steps in
ChIP-exo). We want to apply our QC pipeline to ChIP-Nexus. Depending
on the depth of the regions formed by ChIP-Nexus data, we may need to
modify the QC pipeline in the step where it partitions the reads into
regions, but the $\mbox{local-NSC}$ may be used to prove this
statement.

@@

The data showed in Figure \ref{fig:exo_example} is actually part of a
more interesting experiment from Professor Landick's lab. We have 2
biological replicates of the $\sig$, $\beta'_f$ and $\beta$
transcription factors under two conditions, one where rifampicin was
applied by 20 minutes and another where it wasn't applied. As seen in
the Figure, there is ChIP-exo data for the 3 factors and PE ChIP-Seq
for $\sig$ and $\beta'_f$ (in the Figure both replicates were pooled).

$\sig$ factor is a transcription initiation factor of housekeeping
genes in \emph{E. Coli}, and both $\beta$ factors transcribe the DNA into
mRNA. Hence, we want to use this data to have a better understanding
of the transcription in \emph{E. Coli}.

@@

\begin{figure}[H]
  \centering
 \includegraphics[width = \textwidth]{/p/keles/ChIPexo/volume5/OpenClosedComplexes/figs/filtered/heatmaps_centroid.png}
 \caption{Centered and normalized heatmaps for PE ChIP-Seq peaks of
   $\sig$ and $\beta'_f$ under rif0 and rif20. The regions are
   organized by the centroids clustering of the $\beta'_f$-rif0
   profiles.}
  \label{fig:hm1}
\end{figure}

Figure \ref{fig:hm1} shows heatmaps of the normalized coverages of PE
ChIP-Seq peaks for $\beta'_f$ and $\sig70$ under rif0 and rif20
treatments. Each row represents the same region, under the four
different samples. As expected the signal for both $\sig$ samples is
centered around the middle, for $\beta'_f$-rif20, the majority of the
signal seems to be centered, hence the transcription was interrupted
and for $\beta'_f$-rif0 we can see clusters formed by open regions.

Then, using a set of high quality PE ChIP-Seq peaks and clustering
with the mclust model by Fraley and Raftery, 2007 \cite{mclust} we
were able to label the a set of regions as closed or open
complexes. Hence, we expect that by using ChIP-exo data we may find
different patterns that represent transcription units whose output was
limited at different steps in the formation of EC$_2$.

@@

Last summer, we entered to a contest from ENCODE's functional
characterization group to fit a model to classify enhancers. Roughly
speaking, the dataset consisted of a set of regions
$\mathcal{R}_1,\cdots,\mathcal{R}_n$ and a set of labels
$Y_1,\cdots,Y_n \in \{0,1,\}$. Hence, we built a series of features
$X_1,\cdots,X_p$ and fitted a model. The model was trained by using
10-fold CV and we selected the tunned parameters that maximized the
area under the ROC curve. Then, an independent test set was provided
for which we delivered a set of predictions. However, there are in
total another $m >> n$ regions in the experiment, and the cost of
labeling those $m$ regions may be high (in either the actual monetary
cost of the experiment or the time the scientists spends labeling
those regions). Those $n+m$ regions are H3K27ac peaks, hence they were
ranked and the first $n$ were selected randomly from groups with high,
medium and low signal. This raises the following questions:
